Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2019-07-24 GWLDCI.tex
almost through section 3.2

In \cite{wilsonloop} Agarwala and Amat showed that every admissible Wilson loop diagrams can be identified with a unique positroid cell, and observed that this mapping is neither one-to-one nor onto. In this paper we give a simple characterization of this lack of injectivity in terms of an equivalence relation on the Wilson loop diagrams, and enumerate the number of different Wilson loop diagrams that correspond to each positroid cell. We also study the lack of surjectivity, and show that the number of distinct positroid cells associated to Wilson loop diagrams of a given size is given by counting the directions in an associahedron.

@@

The main result of Section~\ref{sec equiv} is Theorem~\ref{same matroid iff equiv}, which says that two Wilson loop diagrams define the same matroid if and only if they are equivalent.  In subsection~\ref{sec: polygon partitions} we relate exact subdiagrams of Wilson loop diagrams to triangulations of polygons, and hence show that every Wilson loop diagram can be uniquely decomposed as a collection of exact subdiagrams (Corollary~\ref{uniqueproppartitioncor}). In subsection~\ref{sec: exact diagram matroidal props} we examine the matroidal properties of the exact subdiagrams, and prove that a subdiagram of a Wilson loop diagram defines a uniform matroid if and only if the subdiagram is exact (Theorem \ref{exactuniformthm}). This uniform matroid is in fact equal to the dual matroid restricted to the exact subdiagram (Remark \ref{remark exact dual restiction}).  Finally, subsection~\ref{sec: matroids and equivalence} is devoted to the proof of Theorem~\ref{same matroid iff equiv}, along with an easy corollary which gives a formula for the size of each equivalence class (Corollary~\ref{number of equiv diagrams}).

@@

Finally, we describe several important types of matroids. A matroid of rank $k$ with a ground set of size $n$ is called {\em realizable} if there exists some $A$ in the $k\times n$ real Grassmannian $\mathbb{G}_{\mathbb{R}}(k,n)$ whose non-zero $k\times k$ minors are exactly those with columns indexed by an element of $\cB$. If all the maximal minors of the matrix are nonnegative then the matroid realized by this matrix is called a {\em positroid}.  \sanote{we seem to have eliminated all other reference to positroid in this paper. Shoudl we get rid of this? I would prefer to keep this, and say that admissible Wilson loop diagrams define positroids, but leave it there. For me, the point of this paper is to elucidate the map from WLDs to matroids, showing what we already know about the image is part of that.} In other words a positroid is a matroid\footnote{Note that matroid isomorphism allows arbitrary permutations of the ground set, while positroid isomorphism only allows cyclic permutations in order to preserve the nonnegativity of the minors.  In particular, the property of being a positroid is not preserved under matroid isomorphism.} that is realized by an element of the totally nonnegative Grassmannian $\Gr(k,n)$. Finally, a {\em uniform matroid} of rank $r$ is a matroid in which any set of size $\leq r$ is independent.

Matroid theory relates to the study of Wilson loop diagrams as follows. In \cite{wilsonloop}, Agarwala and Amat show that every admissible Wilson loop diagram with $k$ propagators defines a positroid of rank $k$, and that the independent sets can be read directly from the diagram:

@@

This section explores the correspondence between Wilson loop diagrams and positroids that was established in \cite{wilsonloop}. In that paper, the authors show that admissible Wilson loop diagrams define matroids that are also positroids, Theorem \ref{put it back in then ref}. Furthermore, they show that if two Wilson loop diagrams are equivalent, they define the same matroid [ref thm].

In this section we develop this analysis further, showing the converse of Theorem [ref] from loc. cit. (see Theorem \ref{same matroid iff equiv}). We also enumerate the number of diagrams that map to an particular matroid in Corollary \ref{number of equiv diagrams}. To do this, we establish a connection between Wilson loop diagrams and polygon dissections in subsection \ref{sec: polygon partitions}, and uncover some matroidal properties of exact subdiagrams in subsection \ref{sec: exact diagram matroidal props}. This includes the fact that exact subdiagrams correspond to uniform matroids (Theorem \ref{exactuniformthm}), and that the associated matroid of an exact subdiagram can be written as as a contraction by the complementary propagator flat (Theorem \ref{exact diagrams contractions}).\sanote{need to discuss positroids, and whether positroids should appear earlier so that they can be discussed here.}

@@

The key point to note is that the subgraph $t$ depends on how the propagators of $P$ sit inside $[n]$, not on how they sit in \hlfix{the subdiagram $(P,V(P))$, which would have an outer edges between vertices 8 and 1, between 2 and 4, and between 4 and 8, but would not include vertex 3}{Sian replaced $W_P$ with $(P,V(P))$ here and now isn't sure if that was right because this sentence doesn't make sense?}.  In particular, $t$ is not the subgraph of $\tau(W)$ consisting only of edges corresponding to propagators of $P$ (as it may also include some of the outer edges of $\tau(W)$), nor is it the polygon dissection of $(P,V(P))$ viewed as a Wilson loop diagram in its own right (which would have vertex set equal to $V(P)$).
\end{eg}

@@

\begin{lem} \label{maxexactcomplementrank}
Let $W = (\cP, [n])$ be an admissible Wilson loop diagram, and $P \subseteq \cP$. Then: \begin{enumerate}
\item If $(P,V(P))$ is an exact subdiagram of $W$, then $F(P^c)$ is a dependent flat in $W$.
\item If $(P,V(P))$ is a maximal exact subdiagram of $W$, then $F(P^c)$ is a cyclic flat in $W$.
\item If $(P,V(P))$ is an exact subdiagram in $W$, then $\rk(F(P^c)) = |P^c|$.
\end{enumerate} \sanote{end here}
\end{lem}

@@

We know from Lemma~\ref{lem facts about WLD matroids} that $F(P^c)$ is always a flat, so for part (1) we only need to show that $F(P^c)$ is dependent.

Since $W$ is admissible, we have $n \geq |\cP| + 4$. Rewriting this inequality as
\[|V(P)| + |F(P^c)|  \geq  |P| + |P^c| +4,\]
and combining it with the fact that $|V(P)| = |P| + 3$ (from the exactness of $(P,V(P))$), we obtain
\begin{equation}\label{eq F is dependent}|F(P^c)| > |P^c| \;.\end{equation}
Equation \eqref{eq F is dependent} implies that $F(P^c)$ supports fewer propagators than the number of vertices it contains, i.e. $F(P^c)$ is not an independent set.

For part (2), suppose that $(P, V(P))$ is a maximal exact subdiagram. By Corollary \ref{classifyflats} we can decompose $F(P^c)$ as
\begin{equation}\label{eq decompose F}V(P)^c = F(P^c) = C \sqcup S = F(\Prop(C)) \sqcup S,\end{equation}
where $C$ is the largest cyclic flat contained in $F(P^c)$ and $S$ is an  independent set. Note that $C$ must be non-empty since $F(P^c)$ is non-empty and dependent (by part (1)).

@@

Finally, note that since $F(P^c) = C \sqcup S$ (equation~\eqref{eq decompose F}), we have
\[\Prop(C) \subseteq \Prop(F(P^c)) \subseteq P^c.\]
Taking complements, we obtain $P \subseteq \Prop(C)^c = Q$, and hence that $(P, V(P))$ is a subdiagram of $(Q, V(Q))$. Since $(P, V(P))$ is maximally exact by hypothesis, and $(Q, V(Q))$ is exact by equation \eqref{Q defines exact}, we must have $P = Q$. Therefore $S = \emptyset$ and $F(P^c)$ is a cyclic flat.

@@

In \cite{wilsonloop}, the authors show that all admissible Wilson loop diagrams correspond to positroids. That is, they correspond to matroids represented by elements of the positive Grassmannians $\Gr(|\cP|, n)$. Any positroid of rank $k$ on $n$ elements defines a subspace of the positive Grassmannians $\Gr(k, n)$, namely the points which represent it. These subspaces give a CW structure on $\Gr(k,n)$, with each positroid defining a cell.

>>>>

In \cite{wilsonloop} Agarwala and Amat showed that every weakly admissible Wilson loop diagrams can be identified with a type of matroid called a positroid, and observed that this mapping is neither one-to-one nor onto. Each such positroid corresponds to a positroid cell in the appropriate positive Grassmannian, $\Gr(k,n)$ \cite{Postnikov}\sanote{Is this the right citation?}. In this paper we give a simple characterization of this lack of injectivity in terms of an equivalence relation on the Wilson loop diagrams, and enumerate the number of different Wilson loop diagrams that correspond to each positroid cell. We also study the lack of surjectivity, and show that the number of distinct positroid cells associated to Wilson loop diagrams of a given size is given by counting the directions in an associahedron.

@@

The main result of Section~\ref{sec equiv} is Theorem~\ref{same matroid iff equiv}, which says that two Wilson loop diagrams define the same matroid (positroid) if and only if they are equivalent.  In subsection~\ref{sec: polygon partitions} we relate exact subdiagrams of Wilson loop diagrams to triangulations of polygons, and hence show that every Wilson loop diagram can be uniquely decomposed as a collection of exact subdiagrams (Corollary~\ref{uniqueproppartitioncor}). In subsection~\ref{sec: exact diagram matroidal props} we examine the matroidal properties of the exact subdiagrams, and prove that a subdiagram of a Wilson loop diagram defines a uniform matroid if and only if the subdiagram is exact (Theorem \ref{exactuniformthm}). This uniform matroid is in fact equal to the dual matroid restricted to the exact subdiagram (Remark \ref{remark exact dual restiction}).  Finally, subsection~\ref{sec: matroids and equivalence} is devoted to the proof of Theorem~\ref{same matroid iff equiv}, along with an easy corollary which gives a formula for the size of each equivalence class (Corollary~\ref{number of equiv diagrams}).

@@

Finally, we describe several important types of matroids. A matroid of rank $k$ with a ground set of size $n$ is called {\em realizable} if there exists some $A$ in the $k\times n$ real Grassmannian $\mathbb{G}_{\mathbb{R}}(k,n)$ whose non-zero $k\times k$ minors are exactly those with columns indexed by an element of $\cB$. If all the maximal minors of the matrix are nonnegative then the matroid realized by this matrix is called a {\em positroid}.  In other words a positroid is a matroid\footnote{Note that matroid isomorphism allows arbitrary permutations of the ground set, while positroid isomorphism only allows cyclic permutations in order to preserve the nonnegativity of the minors.  In particular, the property of being a positroid is not preserved under matroid isomorphism.} that is realized by an element of the totally nonnegative Grassmannian $\Gr(k,n)$. Finally, a {\em uniform matroid} of rank $r$ is a matroid in which any set of size $\leq r$ is independent.

Matroid theory relates to the study of Wilson loop diagrams as follows. In \cite{wilsonloop}, Agarwala and Amat show that every weakly admissible Wilson loop diagram with $k$ propagators defines a positroid of rank $k$ on a base set of size $n$, called $M(W)$. In other words, by \cite{Postnikov}, each admissible Wilson loop diagram corresponds to a convex subset of $\Gr(k,n)$, thus connecting the geometry of Wilson loop diagrams with the geometry underlying the Amplituhedron. The independent sets of $M(W)$ can be read directly from the diagram:

@@

This section explores the correspondence between Wilson loop diagrams and positroids that was established in \cite{wilsonloop}, (Theorem 3.38). In that paper, the authors show that admissible Wilson loop diagrams define matroids that are also positroids. Furthermore, they show that if two Wilson loop diagrams are equivalent, they define the same matroid(Theorems 1.18 and 3.41).

In this section we develop this analysis further, showing the converse of this result from loc. cit. (see Theorem \ref{same matroid iff equiv}). We also enumerate the number of diagrams that map to an particular matroid in Corollary \ref{number of equiv diagrams}. To do this, we establish a connection between Wilson loop diagrams and polygon dissections in subsection \ref{sec: polygon partitions}, and uncover some matroidal properties of exact subdiagrams in subsection \ref{sec: exact diagram matroidal props}. This includes the fact that exact subdiagrams correspond to uniform matroids (Theorem \ref{exactuniformthm}), and that the associated matroid of an exact subdiagram can be written as as a contraction by the complementary propagator flat (Theorem \ref{exact diagrams contractions}).

@@

The key point to note is that the subgraph $t$ depends on how the propagators of $P$ sit inside $[n]$, not on how they sit in \hlfix{the subdiagram $(P,V(P))$, which would have an outer edges between vertices 9 and 1, and between 5 and 8}{Sian replaced $W_P$ with $(P,V(P))$ here and now isn't sure if that was right because this sentence doesn't make sense?}.  In particular, $t$ is not the subgraph of $\tau(W)$ consisting only of edges corresponding to propagators of $P$ (as it may also include some of the outer edges of $\tau(W)$), nor is it the polygon dissection of $(P,V(P))$ viewed as a Wilson loop diagram in its own right (which would have vertex set equal to $V(P)$).
\end{eg}

@@

\begin{lem} \label{maxexactcomplementrank}
Let $W = (\cP, [n])$ be an admissible Wilson loop diagram, and $P \subseteq \cP$. Then: \begin{enumerate}
\item If $(P,V(P))$ is an exact subdiagram of $W$, then $F(P^c)$ is not an independent flat in $W$.
\item If $(P,V(P))$ is a maximal exact subdiagram of $W$, then $F(P^c)$ is a cyclic flat in $W$.
\item If $(P,V(P))$ is an exact subdiagram in $W$, then $\rk(F(P^c)) = |P^c|$.
\end{enumerate} 
\end{lem}

@@

We know from Lemma~\ref{lem facts about WLD matroids} that $F(P^c)$ is always a flat, so for part (1) we only need to show that $F(P^c)$ is not independent.

Since $W$ is admissible, we have $n \geq |\cP| + 4$. Rewriting this inequality as
\[|V(P)| + |F(P^c)|  \geq  |P| + |P^c| +4,\]
and combining it with the fact that $|V(P)| = |P| + 3$ (from the exactness of $(P,V(P))$), we obtain
\begin{equation}\label{eq F not independent}|F(P^c)| > |P^c| \;.\end{equation}
Equation \eqref{eq F not independent} implies that $F(P^c)$ supports fewer propagators than the number of vertices it contains, i.e. $F(P^c)$ is not an independent set.

For part (2), suppose that $(P, V(P))$ is a maximal exact subdiagram. By Lemma \ref{lem decompose flat} (1) and Corollary \ref{classifyflats} we can decompose $F(P^c)$ as
\begin{equation}\label{eq decompose F}V(P)^c = F(P^c) = C \sqcup S = F(\Prop(C)) \sqcup S,\end{equation}
where $C$ is the largest cyclic flat contained in $F(P^c)$ and $S$ is an  independent set. We show that $\Prop(C) = P^c$, thus forcing $S$ to be empty. Note that $C$ must be non-empty since $F(P^c)$ is non-empty and not independent (by part (1)).

By Lemma \ref{lem facts about WLD matroids}(2), $\rk(C) = |\Prop(C)|$. Since $S$ is independent, Lemma \ref{lem facts about WLD matroids}(1) gives $|S| = \rk(S)$. By similar arguments to Lemma~\ref{lem decompose flat}, the fact that $S$ is independent and that $C$ is the union of all circuits in $F(P^c)$ implies that every element of $S$ is independent of $C$. Combining these give the first two inequalities below
\ba \rk\big(F(P^c)\big) = \rk(C) + \rk(S)  = |\Prop(C)| + |S| \leq |\Prop(F(P^c))|\;,\label{rank of comp flat} \ea % this equation doesn't really need to be displayed, but it was causing a badbox  
while the inequality comes from the bound on ranks of vertex sets.

Furthermore, from Remark \ref{alt F(P) rmk}, we see that $\Prop (F(P^c)) \subseteq P^c$. Therefore equation \eqref{rank of comp flat} gives \bas |\Prop (C)| + |S| \leq |P^c|  \;.\eas For ease of future notation, write $Q = \Prop(C)^c$. 
Rearranging and adding $|P|$ to both sides gives \ba |P| + |S| \leq |P| + |P^c| - |\Prop(C)| = |Q| \;.\label{P+Sbound}\ea Taking complements of and rearranging equation \eqref{eq decompose F} we have
\bas V(P) \sqcup S = C^c = F(\Prop(C))^c  = V(Q)\;. \eas
Combining this with equation \eqref{P+Sbound} gives 
\ba |V(Q)| = | V(P) | + |S| = | P | + 3 + |S| \leq |Q| +3 \;. \label{Q defines exact}\ea
Since $W$ is admissible this cannot be a strict inequality, so we have $|V(Q)| = |Q| +3$ and the subdiagram $(Q, V(Q))$ is exact.

Since \bas \Prop(C) \subseteq \Prop(F(P^c)) \subseteq P^c \eas taking complements gives $P \subseteq \Prop(C)^c = Q$. In other words, $(P, V(P))$ is an exact subdiagram of $(Q, V(Q))$. Since $(P, V(P))$ is a maximal exact subdiagram, $P=Q$, or $\Prop (C) = P^c$, and thus $S = \emptyset$, forcing $F(P^c)$ to be a cyclic flat. \sanote{the previous proof of (2) (commented out) was making my head spin. I think this is more straightforward, butI might just be rearranging deckchairs.}
\begin{comment}For part (2), suppose that $(P, V(P))$ is a maximal exact subdiagram. By Lemma \ref{lem decompose flat} (1) and Corollary \ref{classifyflats} we can decompose $F(P^c)$ as
\begin{equation}\label{eq decompose F}V(P)^c = F(P^c) = C \sqcup S = F(\Prop(C)) \sqcup S,\end{equation}
where $C$ is the largest cyclic flat contained in $F(P^c)$ and $S$ is an  independent set. We show that $\Prop(C) = P^c$, thus forcing $S$ to be empty. Note that $C$ must be non-empty since $F(P^c)$ is non-empty and not independent (by part (1)).

@@

Finally, note that since $F(P^c) = C \sqcup S$ (equation~\eqref{eq decompose F}), we have
\[\Prop(C) \subseteq \Prop(F(P^c)) \subseteq P^c.\]
Taking complements, we obtain $P \subseteq \Prop(C)^c = Q$, and hence that $(P, V(P))$ is a subdiagram of $(Q, V(Q))$. Since $(P, V(P))$ is maximally exact by hypothesis, and $(Q, V(Q))$ is exact by equation \eqref{Q defines exact}, we must have $P = Q$. Therefore $S = \emptyset$ and $F(P^c)$ is a cyclic flat.
\end{comment}

@@

In \cite{wilsonloop}, the authors show that all admissible Wilson loop diagrams correspond to positroids. That is, they correspond to matroids represented by elements of the positive Grassmannians $\Gr(|\cP|, n)$. Any positroid of rank $k$ on $n$ elements defines a subspace of the positive Grassmannians $\Gr(k, n)$, namely the points which represent it \cite{Postnikov}. These subspaces give a CW structure on $\Gr(k,n)$, with each positroid defining a cell.  \sanote{Susama end here}
